# My first simple Rmd file

Don't forget to install knitr package.

```{r, eval=FALSE}
install.packages("knitr")
```

# Introduction

This is an `.Rmd` document.  Maybe you have heard of `Sweave`?
Well, `knitr` is like modern `Sweave`.

* can work with markdown files (simpler) and LaTeX.
* more flexible re: graphics
* caching
* fancy look and feel.

## A simple example

Let's get R to do some simple maths.

```{r}
x <- 1:3
y <- 4:6
outer(x, y)
```
As well as having chunks in separate paragaphs, you can have some
inline computation.  For example, the mean of x is `r mean(x)` and
the sum of x and y is `r x+y`.  Chunks can be named and then
referred to later.

## Plotting is easy too

```{r plot-chunk}
x <- seq(from=0, to=2*pi, length=1000)
y <- cos(3*x)
plot(x, y, type='l', col='blue')
```

## Tables 

See also the [xtable](http://cran.r-project.org/web/packages/xtable/)
package.

```{r,results='asis'}
kable(head(iris[,1:3]), format='html')
```

# Okay, what do you need to know?


1. What is markdown syntax?  [Markdown](http://daringfireball.net/projects/markdown)

2. What can `knitr` do? [Knitr home page](http://yihui.name/knitr/)


# What are vignettes?

Vignettes are self-contained documents that typically describe in
detail how a package works.  See a list of them with `vignette()`.
*knitr* has many vignettes, e.g. `vignette('knitr-refcard')`.

# Installing vignettes into a package.

- add `*.Rmd` files under the `vignettes` directory
- add `VignetteBuilder: knitr` to the `DESCRIPTION` file
- specify the vignette engine `\VignetteEngine{knitr::knitr}` in the
  `Rmd` files (inside HTML comments)

Now build, check and install your package.

# Exercise:

Prepare a `Rmd` document for the `sequences` package. Look at the
original vignette to get started:

```{r, eval=FALSE}
vignette("sequences", package = "sequences")
```

# How to compile this document

This should work cross platform; first, the document is knitted to
make a markdown file; then the "markdown" package converts this to
html which can be rendered.

```{r, eval=FALSE}
library("knitr")
knit2html("simple.Rmd")
if (interactive()) browseURL("simple.html")
```

In R studio, it is even simpler; just hit the special "Knit HTML"
button!

If you prefer the pdf format you need to have a latex compiler, so the
following may not work for you:

```{r, eval=FALSE}
require("knitr")
## Rmd -> md
knit('simple.Rmd')
## Rmd -> html
knit2html('simple.Rmd')
browseURL('simple.html') ## view it

## Rnd -> pdf
knit2pdf("simple.Rmd")
```

## About this document

```{r}
sessionInfo()
```
